---
title: "Untitled"
author: "Mohammed Faizan"
date: '2022-09-19'
output: html_document
---

```{r setup,echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  echo = FALSE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  out.width = "100%",
  fig.align = "center",
  fig.retina = 4,
  cache = TRUE
)
```

```{r}
library(ggrepel)
library(tidyverse)
library(tidymodels)
library(discrim)
library(mgcv)
library(broom)
library(patchwork)
library(kableExtra)
library(viridisLite)
library(plotly)
library(tourr)

library(discrim)
library(parsnip)
library(MASS)
library(tidymodels)
library(rsample)
library(yardstick)
library(spinifex)
library(GGally)
library(palmerpenguins)
library(fitzRoy)
library(boot)


```

```{r writing_packages_bibliographies3}
knitr::write_bib(c(.packages()), "packages.bib")
```


# Introduction and motivation

# Data description

```{r }
us_ineq0 <- read_csv(here::here("Inequality_data.csv")) %>% 
  rename("state"="State")

us_ineq_long0 <- us_ineq0 %>% 
  pivot_longer(cols = c(2:12),
               names_to = "year",
               values_to = "gini")

us_ineq <- read_csv(here::here("Inequality_data.csv")) %>% 
  rename("state"="State") %>% 
  mutate(`1968`=case_when(state=="Oregon"~round(0.468620508909225,7),
                          TRUE~`1968`))

us_ineq_long <- us_ineq %>% 
  pivot_longer(cols = c(2:12),
               names_to = "year",
               values_to = "gini")

```

#Exploratory Data Analysis

```{r ginidis, fig.cap="Density of Gini"}
#votes distribution, bimodal. mostly around 0.5
p1 <- us_ineq_long %>% 
  filter(state !="United States") %>%
  ggplot(mapping = aes(x=gini)) +
  geom_histogram(binwidth=0.05,
                 fill = "#beaed4")+
  theme_bw() +
  labs(x = "Gini") +
  coord_cartesian(xlim = c(0,1))

#US ts
p2 <- us_ineq_long %>% 
  filter(state =="United States") %>%
  ungroup() %>% 
  ggplot(mapping = aes(x=year,y=gini)) +
  geom_col(size=3,fill = "#beaed4")+
  theme_bw() +
  coord_cartesian(ylim = c(0,1))


```
```{r eval=FALSE}
#bar distribution
#rating distribution
us_ineq_long %>% 
  mutate(lsgt50 = case_when(
    gini>0.5 ~ "> 0.5",
    gini<0.5 ~ "< 0.5",
    TRUE ~ "0.5")) %>% 
  mutate(lsgt36 = case_when(
    gini>0.5 & gini<=0.625 ~ "0.625",
    gini<=0.5 & gini>=0.375 ~ "0.375",
    TRUE ~ "other")) %>% 
  ggplot(mapping = aes(x=lsgt50,y=gini, fill=lsgt36),position="stack") +
  geom_col() +
  theme_bw() +
  scale_fill_brewer("Gini",
                    palette = "Accent") +
  labs(x = "High/Low",
       y = "Gini")

#fewer states with lower inequality. Percentage of other in higher gini is higher than in lower gini. This means that more states have higher inequality than. Q4>Q1

```
The Gini Index, or Gini coefficient, is a statistical measure of distribution that is often used to track economic inequality. It measures how wealth is distributed in a given population. The output is a value between 0 and 1. In Oregon and throughout the country, Gini Index is a major focus of the American Community Survey (ACS), conducted each year by the U.S. Census Bureau.

Higher values mean greater inequality. A Gini value of 0 means a perfectly equal society, where everyone's income is the same. A value of 1, on the other hand, represents perfect inequality—a society in which only one person or group has all the wealth. This report explores how Oregon’s Gini Index is changing—and how the data compares to the United States as a whole.
```{r}

#Oregon 1968 1000.0000000 wrong value must be corrected to 0.468620508909225.
# https://www.livestories.com/statistics/oregon/gini-index-income-inequality

p3 <- us_ineq_long0 %>% 
  filter(state !="United States") %>%
  ggplot(mapping = aes(x=year,y=gini)) +
  geom_line(colour = "#beaed4")+
  theme_bw() +
  coord_cartesian(ylim = c(0,1000))
```

```{r}
p4 <- us_ineq_long %>% 
  filter(state !="United States") %>%
  ggplot(mapping = aes(x=year,y=gini,color="state")) +
  geom_line(colour = "#beaed4")+
  theme_bw() 
```

# Analysis and findings


# Conducting and interpreting a PCA

# MDS

# Conclusion

Principal component analysis is often used to create indicator variables.  



```{r echo=TRUE}
# duplicate playernames.
# aflw_num1 %>% 
#   distinct(name,player_player_id) %>% 
#   count(player_player_id) %>% 
#   arrange(-n)

#length(unique(aflw$player_player_id)) 370


aflw_num1 %>% filter(player_player_id %in% c("CD_I1009838","CD_I1011127")) %>%
  arrange(player_player_id) %>% view()

aflw_num2 <-  aflw_num1 %>% 
  mutate(name =case_when(
    str_detect(name, "JesseTawhiao-Wardlaw") ~ "JesseWardlaw",
    str_detect(name, "JessicaFoley") ~ "JessFoley",
    TRUE ~ name))
#only numerics
aflw_num3 <- aflw_num2 %>% 
  dplyr::select(-c(1:3)) %>% 
  dplyr::select(-c(35:40))
  
aflw_num4 <- aflw_num3 %>% cbind(aflw_num2 %>% dplyr::select(name))
#aflw_num4[1:34]
#length(unique(aflw_num2$name)) 370
```

a. (1pt) Compute the total variance assuming standardised variables.

```{r}
# tss_ns <- apply(aflw_num3, 2, var) #not standardised
# sum(tss_ns)

afl_pc <- prcomp(aflw_num4[1:34], scale = TRUE, center = TRUE)
#afl_pc$sdev
tss_s <- sum(afl_pc$sdev^2) #standardised
```


**the total variance assuming standardised variables in the data is `r tss_s`.**

b. (1pt) Compute a PCA. What proportion of variance do the first two PCs explain?

```{r}
screeplot(afl_pc, type="l")

pve <- afl_pc$sdev^2 / tss_s

#pve[1]+pve[2] #38.66%
```


**First two PCs explain about `r round((pve[1]+pve[2])*100,2)`% of variance.**

c. (1pt) Examine the loadings for the PC1 and PC2. Make plots of the loadings (like done in lecture and during tutorial). 

```{r}
#summary(afl_pc) 

#afl_pc$rotation[,1:2]

afl_pc_loadings <- as_tibble(afl_pc$rotation[,1:2]) %>%
  mutate(variable = rownames(afl_pc$rotation), 
         indx = 1:nrow(afl_pc$rotation),
         ymin=rep(0, nrow(afl_pc$rotation)))
ggplot(afl_pc_loadings) + 
  geom_hline(yintercept=c(-1/sqrt(nrow(afl_pc$rotation)),
                          1/sqrt(nrow(afl_pc$rotation))), colour="red") + 
  geom_errorbar(aes(x=indx, ymin=ymin, ymax=PC1)) +
  geom_point(aes(x=indx, y=PC1))


ggplot(afl_pc_loadings) + 
  geom_hline(yintercept=c(-1/sqrt(nrow(afl_pc$rotation)),
                          1/sqrt(nrow(afl_pc$rotation))), colour="red") + 
  geom_errorbar(aes(x=variable, ymin=ymin, ymax=PC2)) +
  geom_point(aes(x=variable, y=PC2)) +
  theme(axis.text.x = element_text(angle = 45))
```
```{r loadings}

afl_pc_loadings %>% arrange(PC1) %>% view()
ggplot(afl_pc_loadings,aes(PC1,PC2)) +
  geom_point() +
  geom_text(aes(label=variable)) +
  labs(title="Loadings")
```



```{r scores}
afl_pc_scores <- as_tibble(afl_pc$x[,1:2]) %>% 
  cbind(aflw_num2 %>% dplyr::select(name))

afl_pc_scores %>% ggplot(aes(x=PC1,y=PC2)) +
  geom_point() +
  theme(legend.position = "none") +
  labs(title="Scores")
```

**Almost all loadings for PC1 are negative or on one side of zero. The loadings for PC2 are present on both sides of zero indicating a contrast of inference for the high and low values on PC2.**

d. (1pt) Explain why PC2 might be used as an indicator for offensive player skills (in 30 words or less). 

**The significant loadings for PC2 are skills related to offensive(eg, goal, goal accuracy, behinds etc) and defensive(intercepts and rebounds). These contrasting skills are present on opposite sides of zero proving that PC2 can be used as a indicator for offensive player skills.**

e. (1pt) Use bootstrap to assess which variables could be considered unimportant for PC2 (ie loadings not significantly different from 0).

# Loadings for PC2

```{r out.width="60%", fig.width=6, fig.height=4}
library(boot)
compute_PC2 <- function(data, index) {
  pc2 <- prcomp(data[index,], center=TRUE, scale=TRUE)$rotation[,2]
  # Coordinate signs
  if (sign(pc2[1]) < 0) 
    pc2 <- -pc2 
  return(pc2)
}
# Make sure sign of first PC element is positive
PC2_boot <- boot(data=aflw_num4[,1:34], compute_PC2, R=1000)
colnames(PC2_boot$t) <- colnames(aflw_num4[,1:34])
PC2_boot_ci <- as_tibble(PC2_boot$t) %>%
  gather(var, coef) %>% 
  mutate(var = factor(var, levels=colnames(aflw_num4[,1:34]))) %>%
  group_by(var) %>%
  summarise(q2.5 = quantile(coef, 0.025), 
            q5 = median(coef),
            q97.5 = quantile(coef, 0.975)) %>%
  mutate(t0 = PC2_boot$t0) 
ggplot(PC2_boot_ci, aes(x=var, y=t0)) + 
  geom_hline(yintercept=0, size=3, colour="white") +
  geom_hline(yintercept=c(1/sqrt(nrow(afl_pc$rotation)), -1/sqrt(nrow(afl_pc$rotation))), linetype=2, colour="red") +
  geom_point() +
  geom_errorbar(aes(ymin=q2.5, ymax=q97.5), width=0.1) +
  xlab("") + ylab("coefficient") +
  theme(axis.text.x = element_text(angle = 80))
``` 

**The variables which have zero in their error band in the above plot are unimportant for PC2.**

f. (1pt) Write down the formula for your new indicator variable. Explain your reasoning (in 30 words or less). 

```{r}

loadings <- tibble(var=names(afl_pc$rotation[,2]),
                   loadingPC2=afl_pc$rotation[,2])
loadings %>% 
  filter(loadingPC2>1/sqrt(nrow(afl_pc$rotation))) %>% 
  rbind(loadings %>% 
  filter(loadingPC2< -1/sqrt(nrow(afl_pc$rotation))) ) %>% 
  kable(caption = "Loading Vector with Important Variables")
   
                        
```

$PC2=goals*0.38+behinds*0.27+marks_inside50*0.33+goal_accuracy*0.30+tackles_inside50*0.2+shots_at_goal*0.42+score_involvements*0.3+rebound50s*-0.2+intercepts*-0.24$.

**Since the goal of PC2 is to indicate offfensive skills all other skills irrelavant to this task are not important and are hence emoved from the formula for PC2.**

g. (2pt) Make a biplot of the first two PCs. Which few players have the highest values, and which few players have the lowest value on PC2? What does it mean for a player to have a high value on PC2 (in 50 words or less)? 


```{r eval=FALSE}
biplot(afl_pc, scale = 0) +geom_text(aes(label=variable))
```

```{r}
library(ggrepel)
afl_pca_var <- tibble(n=1:length(afl_pc$sdev), evl=afl_pc$sdev^2)
afl_pca_pcs <- as_tibble(afl_pc$x[,1:2]) %>%
  mutate(name=aflw_num4$name) %>% 
  mutate(club=aflw_num2$team_name)
afl_pca_evc <- as_tibble(afl_pc$rotation[,1:2]) %>% 
  mutate(origin=rep(0, ncol(aflw_num4)-1), variable=colnames(aflw_num4)[1:34],
         varname=rownames(afl_pc$rotation)) %>%
  mutate(PC1s = PC1*(afl_pca_var$evl[1]*2.5), 
         PC2s = PC2*(afl_pca_var$evl[2]*2.5))
pca_p <- ggplot() + 
  geom_segment(data=afl_pca_evc %>% filter(varname %in% names(loadings)) , aes(x=origin, xend=PC1s, y=origin, yend=PC2s), colour="orange") +
  geom_text_repel(data=afl_pca_evc, aes(x=PC1s, y=PC2s, label=variable, nudge_y=sign(PC2)*0.1), colour="orange", nudge_x=0.1) +
  geom_point(data=afl_pca_pcs, aes(x=PC1, y=PC2),alpha=0.1) +
  geom_text_repel(data=filter(afl_pca_pcs, PC2>10), aes(x=PC1, y=PC2, label=name), nudge_y=0.1, nudge_x=-0.1) +
  geom_text_repel(data=filter(afl_pca_pcs, PC2< -3.5), aes(x=PC1, y=PC2, label=name), nudge_y=0.1, nudge_x=-0.1) +
  #geom_text(data=filter(afl_pca_pcs,name=="MadisonPrespakis"), aes(x=PC1, y=PC2, label=name), nudge_y=0.1, nudge_x=-0.1) +
  xlab("PC1") + ylab("PC2") +
  theme(aspect.ratio=1)

filter(afl_pca_pcs, PC2< -3) %>% view()

#high scores on PC2 are more offensive. More goals in the season
# low scores have the best and fairest awards in the 2020 season
pca_p
```

- Players with high PC2 values are:
CoraStaunton, ChloeMolloy, SabreenaDuffy, and SarahD'Arcy
` knitr::combine_words(filter(afl_pca_pcs, PC2>10)$name)`

- Players with high PC2 values are:
PhoebeMonahan, MillieBrown, StaceyLivingstone, AlexandraAnderson, and DanaHooker
` knitr::combine_words(filter(afl_pca_pcs, PC2< -3.5)$name)`


**A high value on PC2 represents the player is more offensive. high scores on PC2 are more offensive. Such players have received leading goalkicker [awards](https://en.wikipedia.org/wiki/2020_AFL_Women%27s_season#Awards) in the season. Players with low PC2 scores are defensive and have received the best and fairest awards in the 2020 season.**

# References



## R packages

```{r, load_refs, echo=FALSE}
library(RefManageR)
bib <- ReadBib(here::here("packages.bib"), check = FALSE)
ui <- "- "
```

```{r, print_refs, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
writeLines(ui)
print(bib, 
  .opts = list(check.entries = FALSE, 
               style = "html", 
               bib.style = "authoryear"))
```
